---
title: "ML - Assignment 2 - Pump it Up"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(plyr)
library(caret)
library(ggplot2)
library(DataExplorer)
library(googleVis)
library(randomForest)
library(lubridate)
library(geosphere)
```

### Import Data
```{r}
setwd("~/Desktop/IE/Semester 2/ML/Assignments/Assignment2")

train <- read.csv("train.csv")
test <- read.csv("test.csv")
labels <- read.csv("train_labels.csv")

train <- merge(train, labels, by = "id", all.x = T)
```

# 1. Exploration - Train Data

```{r}
str(train)
```
Let's transform `region_code` and `district_code` to `factors`. Even though those are numeric variables, they account as factor since each code is related to the `region` variable or a geographic location. 
```{r}
train$region_code <- as.factor(train$region_code)
train$district_code <- as.factor(train$district_code)
```

```{r}
# divide continuous and discrete features
# split_columns (data explorer pkg) creates two data frames, one with continuous and one with discrete values 
values <- split_columns(train)
cat_var <- values$discrete
num_var <- values$continuous
names(cat_var)
names(num_var)
```
## 1.1 Categorical variables
```{r}
names(cat_var)
```

```{r}
# bar plots for discrete features
plot_bar(train)
```
  
The plots above show that different features describe the same thing, but at different level of detail while other features represent a very big number of levels. We analyze those features in groups and see if they're relevant or alternatively if they can be modified or excluded.  
Below are the *groups* of features identified:    
  
* `management_group`, `management`  
* `scheme_management`, `scheme_name`  
* `payment_type`, `payment`  
* `quality_group`, `water_quality`  
* `quantity_group`, `quantity`  
* `source`, `source_type`, `source_class`  
* `waterpoint_type`, `waterpoint_type_group`  
* `region`, `region_code`, `district_code`  
   
## 1.2 Numeric variables
```{r}
names(num_var)
```

```{r}
# box plot for continuous features
plot_boxplot(train[,-1], by = "status_group")
```

```{r}
# hitograms for continuous variables
plot_histogram(train[,-1])
```
The `boxplot` and `histograms` show that many variables have value `0`, being - in most cases - the equivalent of `missing values`.  
Let's see how many and in what percentage, there are for each feature.  
```{r}
# number of "zeros"
print("Number of zeros (0) per feature")
sort(sapply(num_var[, -which(names(num_var)%in%c("id"))], function(x) sum(x == 0)), decreasing = T)
# percentage of "zeros"
print("Percentage of zeros (0) per feature")
sort(sapply(num_var[, -which(names(num_var)%in%c("id"))], function(x) paste(round(sum(x == 0)/length(x),5)*100, "%")), decreasing = T)
```

We exclude:  
  
* `num_private` because about 99% of the values are missing  
* `amount_tsh` because about 70% of the values are missing  
  
```{r}
# remove num_private
train$num_private <- NULL
test$num_private <- NULL

# remove amount_tsh
train$amount_tsh <- NULL
test$amount_tsh <- NULL
```

Other variables have to be treated accordingly to replace `zero` values with appropriate values:  
  
* `population`  
* `construction_year`  
* `gps_height`  
* `longitude` and `latitude`  
  
```{r}
# # number of pumps in each functional status group
# table(train$status_group)
# 
# # as proportions
# prop.table(table(train$status_group))
#   
# # quantity variable vs the status of the pumps
# table(train$quantity, train$status_group)
# 
# # as proportions
# prop.table(table(train$quantity, train$status_group), margin = 1)
```

### Latitude & Longitude

```{r}
# scatter plot: latitude vs longitude with color as status_group
ggplot(subset(train, latitude < 0 & longitude > 0),
    aes(x = latitude, y = longitude, color = status_group)) + 
    geom_point(shape = 1) + 
    theme(legend.position = "top") +
    scale_fill_manual(values=colors)
```



# 2. Feature Engineering

## 2.1 Categorical Variables
We are going to treat the *bundles* of categorical variables described in the section before (`1.1`). For each bundle of varaibles we group by every high level detail feature and analyze the different levels of grain between the variables and if it is possible to improve them.
  
### quantity_group, quantity
In both features there are the same values, but with different naming. We will delete one of them: `quantity`. 
```{r}
train %>% group_by(quantity_group, quantity) %>% tally()
```

```{r}
train$quantity_group <- NULL
test$quantity_group <- NULL
```

### management_group, management
I keep both management and management_group describe the same thing, at different level of detail. Both can be important.  
```{r}
train %>% group_by(management_group, management) %>% tally()
```

### scheme_management, scheme_name
`scheme_name` has too many levels, we remove it.

```{r}
length(levels(train$scheme_name))
length(levels(train$scheme_management))
```

```{r}
train$scheme_name <- NULL
test$scheme_name <- NULL
```

### payment_type, payment
In both features there are the same values, but with different naming. We will delete one of them: `payment`. 
```{r}
train %>% group_by(payment_type, payment) %>% tally()
```

```{r}
train$payment <- NULL
test$payment <- NULL
```

### extraction_type_class, extraction_type_group, extraction_type
The three features concerning `extraction` are related to each other. The `extraction_type` is thus used to create groups based on `extraction_type_group` and other values of `extraction_type`.  
  
```{r}
train %>%
  group_by(extraction_type_class, extraction_type_group, extraction_type) %>% tally()
```

```{r}
train$extraction_type <- as.character(train$extraction_type)
train$extraction_type[train$extraction_type %in% c("cemo", "climax")] <- "other motorpump"
train$extraction_type[train$extraction_type %in% c("other - mkulima/shinyanga", "other - play pump", "walimi")] <- "other handpump"
train$extraction_type[train$extraction_type %in% c("other - swn 81", "swn 80")] <- "swn"
train$extraction_type[train$extraction_type %in% c("india mark ii", "india mark iii")] <- "india mark"

train$extraction_type <- as.factor(train$extraction_type)
# remove extraction_type_group
train$extraction_type_group <- NULL
```

```{r}
test$extraction_type <- as.character(test$extraction_type)

# apply to test data
test$extraction_type[test$extraction_type %in% c("cemo", "climax")] <- "other motorpump"
test$extraction_type[test$extraction_type %in% c("other - mkulima/shinyanga", "other - play pump", "walimi")] <- "other handpump"
test$extraction_type[test$extraction_type %in% c("other - swn 81", "swn 80")] <- "swn"
test$extraction_type[test$extraction_type %in% c("india mark ii", "india mark iii")] <- "india mark"

test$extraction_type <- as.factor(test$extraction_type)

# remove extraction_type_group
test$extraction_type_group <- NULL
```

### quality_group, water_quality
Most of the values are the same, except for some that are better identified in `water_quality` rather than in `quality_group`. We keep the variable with higher level of detail: `water_quality`.  
```{r}
train %>% group_by(quality_group, water_quality) %>% tally()
```

```{r}
train$quality_group <- NULL
test$quality_group <- NULL
```

### source, source_type, source_class
`source_type` and `source` have the same values, but `source` has a higher level of detail. Hence we will delete `source_type` and keep the other two features. Moreover we relabel `unknow` values in `source` with the value `other`.  
```{r}
train %>% group_by(source_class, source_type, source) %>% tally()
```

```{r}
train$source_type <- NULL
test$source_type <- NULL
```

### waterpoint_type, waterpoint_type_group
The features are almost identical. We keep, as before, the feature with higher level of grain: `waterpoint_type`
```{r}
train %>% group_by(waterpoint_type_group, waterpoint_type) %>% tally()
```

```{r}
train$waterpoint_type_group <- NULL
test$waterpoint_type_group <- NULL
```

### installer

There are a lot of terms that are likely the same installer, but have different names. Such as `Government` and `Govt`. We aggregate them taking the first 3 letters of each factor and making them lower case. Then, we can aggregate the most frequent and we put all other variables under `other`.  
  
```{r}
summary(train$installer)[1:15]
#length(unique(train$installer))

# take first 3 letters and make them lowercase
train$installer_grouped <- substr(tolower(train$installer),1,3)
train$installer_grouped[train$installer_grouped %in% c(" ", "", "0", "_", "-")] <- "other"

# top 15 occurrences
install_top_15 <- names(summary(as.factor(train$installer_grouped)))[1:15]
train$installer_grouped[!(train$installer_grouped %in% install_top_15)] <- "other"
train$installer_grouped <- as.factor(train$installer_grouped)

# Apply to test data
test$installer_grouped <- substr(tolower(test$installer),1,3)
test$installer_grouped[test$installer_grouped %in% c(" ", "", "0", "_", "-")] <- "other"
test$installer_grouped[!(test$installer_grouped %in% install_top_15)] <- "other"
test$installer_grouped <- as.factor(test$installer_grouped)
```

## 2.2 Numeric Variables
We tried to replace incorrect (value = `0`) `longitude`, `latitude` and `gps_height` values with the mean of the `region` and related `district_code`. Running the model with the transformation the score got worse, hence we removed it.  

### region, region_code, district_code
We delete some variables related to the position of the pump, since we won't need all of them in the model. 

```{r}
train$region_code <- NULL
test$region_code <- NULL

train$district_code <- NULL
test$district_code <- NULL

train$ward <- NULL
test$ward <- NULL

train$subvillage <- NULL
test$subvillage <- NULL
```


### day, month, distance
We create two variables related to the date:  
  
* `dr_days`: days from the `date_recorded`  
* `dr_month`: the month of `date_recorded`    
  
```{r}
# create days, month and distance variables 
train$dr_days <- as.numeric(as.Date("2014-01-01") - as.Date(train$date_recorded))
train$dr_month <- month(ymd(train$date_recorded))

# apply to test data
test$dr_days <- as.numeric(as.Date("2014-01-01") - as.Date(test$date_recorded))
test$dr_month <- month(ymd(test$date_recorded))
```

We create one variable related to the location:  
  
* `distance`: distance between c(0,0) and c(longitude,latitude) of each pump  
  
```{r}
train$distance <- distGeo(as.matrix(train[,c('longitude','latitude')]), c(0,0))
test$distance <- distGeo(as.matrix(test[,c('longitude','latitude')]), c(0,0))
```

# 3. Prediction
We mainly spent time tuning the model, changing the number of trees `ntree` and the `features`.  
The remaining data set has `r ncol(train)` features. After different attempts with different combination of features, we obtained the best result with the variables in the model below.  

### Random Forest
```{r}
#  Set seed and create a random forest classifier
 set.seed(1234)
 
# model
model_forest <- randomForest(status_group ~ installer_grouped + latitude + longitude + gps_height + scheme_management + public_meeting + permit + water_quality + quantity + payment_type + source_class + source +  management  +  management_group + basin + extraction_type + waterpoint_type + dr_days + dr_month + distance + region + population,
                              data = train, 
                              importance = T, 
                              ntree = 1001, 
                              mtries = 10,
                              nodesize = 2
                              )
 
# predict the values in train
pred_forest_train <- predict(model_forest, train)
```
  
### Evaluation
```{r}
# confusion matrix
confusionMatrix(pred_forest_train, train$status_group)
```

```{r}
# variable importance
importance(model_forest)
varImpPlot(model_forest)
```

# 4. Submission

```{r}
# use same levels
common <- intersect(names(train), names(test)) 
for (p in common) {
  if (class(train[[p]]) == "factor") { 
    levels(test[[p]]) <- levels(train[[p]]) 
   } 
}
 
# Predict using the test data
pred_forest_test <- predict(model_forest, test)
 
## create submission data frame
# submission <- data.frame(test$id)
# submission$status_group <- pred_forest_test
# names(submission)[1] <- "id"
# submission[sort(submission$id),]
# 
# write.csv(submission, file = "/filename/submission.csv", row.names = F)
```

